WIP: Transparent Standby for Low-Power, Resource-Constrained Embedded Systems

### 2018-02-17 LCTES/body.tex
(>) standby infrastructure

Suppose we now want that, at any time, receiving a message via radio should
immediately abort the loop in Figure~\ref{lst.direct}.a.
%
Since the message might arrive concurrently with any of the blocking
operations, we need to change the structure of the program.
%
Figure~\ref{lst.inversion}.a changes the blocking operation \code{delay} to the
polling operation \code{millis}, which immediately returns the number of
milliseconds since the reset.
Now, we start by registering the current time (ln. 1--2) and, on each loop
iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
Since these operations are non-blocking, we can intercalate the execution with
checks for message arrivals (ln. 4--6).
If the time is up, we start counting it again (ln. 10) before proceeding to the
original operations in sequence (ln. 11--13).
%
The original structured style has been drastically violated to accommodate
concurrency.
In the example, we only adapted the \code{delay} operation, but the other
blocking operations (\code{analogRead} and \code{radioWrite}) would also need
to be changed to achieve maximum concurrency.
%
Alternatively, we could resort to ISRs or implement an event-driven
scheduler to handle the operations~\cite{tos.nesc}, but ultimately, the
program readability would still be compromised.

@@

\section{Transparent Standby Mechanisms}
\label{sec.standby}

@@

Figure~\ref{lst.adc} shows the driver for the A/D converter in \CEU.
For simplicity, we assume in the paper that the converter has a single channel
to avoid having to deal with multiplexing.
%
The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
port manipulation in the microcontroller.
The output event (ln. 9--11)

@@

output void ANALOG_REQUEST;    // low-level request
input  int  ANALOG_DONE;       // low-level response
code AnalogRead (void) -> int; // high-level abstraction

@@

output void ANALOG_REQUEST do
  {
    ADMUX = 0x40 | (A0 & 0x07);
    bitSet(ADCSRA, ADIE); // enables interrupt
    bitSet(ADCSRA, ADSC); // starts the conversion
  }
end

async/isr ADC_vect_num do
  { bitClear(ADCSRA, ADIE); } // disables interrupt
  var int value = {ADC}; // register with the value read
  emit ANALOG_DONE(value);
end

code AnalogRead (void) -> int do
  {PM_SET(PM_ANALOG, 1);}
  do finalize with
      {PM_SET(PM_ANALOG, 0);}
  end

  emit ANALOG_REQUEST(pin);
  var int value = await ANALOG_DONE;

@@

void pm_sleep (void) {
  if (PM_GET(PM_TIMER1) ||
      PM_GET(PM_USART)  ||
      PM_GET(PM_SPI))
  {
      LowPower.idle(PM_GET(PM_ADC),...)
    } else if (PM_GET(PM_ADC)) {
      LowPower.adcNoiseReduction(...);
    } else {
      LowPower.powerDown(...);
    }
  }
}
\end{lstlisting}
%\rule{8.4cm}{0.37pt}
\caption{ Power management module for the ATmega328p microcontroller.
\label{lst.pm}
}
\end{figure}

>>>>

Suppose we now want that, at any time, receiving a message via radio should
immediately abort the loop in Figure~\ref{lst.direct}.a.
%
Since the message might arrive concurrently with any of the blocking
operations, we need to change the structure of the program.
%
Figure~\ref{lst.inversion}.a changes the blocking operation \code{delay} to the
polling operation \code{millis}, which immediately returns the number of
milliseconds since the reset.
Now, we start by registering the current time (ln. 1--2) and, on each loop
iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
Since these operations are non-blocking, we can intercalate the execution with
checks for message arrivals (ln. 4--6).
If the time is up, we start counting it again (ln. 10) before proceeding to the
original operations in sequence (ln. 11--13).
%
The original structured style has been drastically violated to accommodate
concurrency.
In the example, we only adapted the \code{delay} operation, but the other
blocking operations (\code{analogRead} and \code{radioWrite}) would also need
to be changed to achieve maximum concurrency.
%
Alternatively, we could resort to ISRs or implement an event-driven
scheduler to handle the operations~\cite{wsn.nesc}, but ultimately, the
program readability would still be compromised.

@@

\section{Standby Infrastructure}
\label{sec.standby}

@@

Figure~\ref{lst.adc} shows the driver for the A/D converter in \CEU.
This code is specific to the \emph{ATmega328p} microcontroller and must be
adapted to work in other platforms.
For simplicity, we assume in the paper that the converter has a single channel
to avoid having to deal with multiplexing.

The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
port manipulation in the microcontroller.
Output events are triggered with the \code{emit} keyword (ln. 29), while input
events are captured with the \code{await} keyword (ln. 30).
%
The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
starts an analog-to-digital conversion in the single channel \code{A0}.
Any code in between \code{\{} and \code{\}} is treated as an inline $C$ chunk,
allowing for easy integration with $C$ for low-level operations.

The \code{async/isr} construct of \CEU defines an ISR which executes
asynchronously with the program when the specified interrupt occurs.
Only ISRs can emit input events to the program.
In the example, we define an ISR to handle \code{ADC} interrupts which fire
whenever a conversion is complete (ln. 17--21).
%
Although the ISR block executes asynchronously on interrupts, the input
emission (ln. 20) only takes effect on a subsequent reaction, when the
synchronous part of the program becomes idle.
%
This way, race conditions are only possible within \code{async/isr} blocks,
which are typically hidden inside device drivers.
\CEU also provides an \code{atomic} primitive to protect critical sections of
code.

The low-level events are the pieces that vary among platforms.
A driver can also offer a higher-level portable abstraction to be exposed to
client code.
%
In the example, the \code{AnalogRead} abstraction (ln. 23--33) takes care of
starting and awaiting the conversion (ln. 29--30), as well as dealing with the
power management runtime (PMR).
%
The \code{PM\_SET(PM\_ADC,1)} (ln. 24) tells the system that, when entering in
sleep mode, the \code{ADC} must be kept running.
%
The \code{PM\_SET(PM\_ADC,0)} inside the \code{finalize} clause (ln. 25--27)
releases the \code{ADC} subsystem.

The \code{finalize} construct of \CEU executes the nested code whenever the
enclosing block terminates or is aborted externally.
%
The example of Figure~\ref{lst.inversion}.b invokes the \code{AnalogRead}
abstraction (ln. 7) concurrently with \code{RadioAvail} (ln. 2).
%
It is possible that the \code{AnalogRead} terminates normally or that a radio
message arrives during the A/D conversion, which causes the \code{AnalogRead}
to abort abruptly.
In either case, the \code{finalize} clause executes and puts the PMR in a
consistent state.

Client code for the driver

 chunk,

@@

output void ADC_REQUEST; // low-level request
input  int  ADC_DONE;    // low-level response
code AnalogRead (void) -> int; // high-level abstraction

@@

output void ADC_REQUEST do
  {
    ADMUX = 0x40 | (A0 & 0x07);
    bitSet(ADCSRA, ADIE); // enables interrupt
    bitSet(ADCSRA, ADSC); // starts the conversion
  }
end

async/isr {ADC_vect_num} do
  { bitClear(ADCSRA, ADIE); } // disables interrupt
  var int value = {ADC}; // reads register with the value
  emit ADC_DONE(value);
end

code AnalogRead (void) -> int do
  {PM_SET(PM_ADC, 1);}
  do finalize with
      {PM_SET(PM_ADC, 0);}
  end

  emit ADC_REQUEST(pin);
  var int value = await ADC_DONE;

@@

void pm_sleep (void) {
  if (PM_GET(PM_TIMER1) ||
      PM_GET(PM_USART)  ||
      PM_GET(PM_SPI))
  {
      LowPower.idle(PM_GET(PM_ADC),...)
    } else if (PM_GET(PM_ADC)) {
      LowPower.adcNoiseReduction(...);
    } else {
      LowPower.powerDown(...);
    }
  }
}
\end{lstlisting}
%\rule{8.4cm}{0.37pt}
\caption{ Power management module for the \emph{ATmega328p} microcontroller.
\label{lst.pm}
}
\end{figure}
